---
title: "Understanding Female Labor Force Participation Rate (LFPR)"
author: "Liping Wang"
date: "12/4/2019"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, include=FALSE}
library(tidyverse)
library(ggplot2)
library(lubridate)
library(wbstats)
library(gridExtra)
library(caret)
library(recipes)
```

```{r, include=FALSE}
wb_data <- wb(indicator = c("SL.TLF.CACT.FE.ZS","NY.GDP.PCAP.CD",
                            "SE.ADT.LITR.FE.ZS","SP.DYN.TFRT.IN",
                            "SG.TIM.UWRK.FE", "SE.COM.DURS",
                            "SE.SEC.CUAT.LO.FE.ZS", 
                            "SE.TER.CUAT.BA.FE.ZS", 
                            "SE.XPD.TOTL.GB.ZS"), 
              startdate = 2005,
              enddate = 2018,
              country = "countries_only") %>% as_tibble()

wb_data <- wb_data %>% mutate(year=as.numeric(date)) %>%
  select(iso3c,year,value,indicatorID,country) %>%
  spread(key = indicatorID, value = "value")

wb_data <- wb_data %>% select(country=country,ccode=iso3c,year=year,
                              f_lfpr=SL.TLF.CACT.FE.ZS,
                              gdp_pcap=NY.GDP.PCAP.CD,
                              f_literacy_rate=SE.ADT.LITR.FE.ZS,
                              com_edu=SE.COM.DURS,
                              f_lower_sec_edu=SE.SEC.CUAT.LO.FE.ZS,
                              f_ba_edu=SE.TER.CUAT.BA.FE.ZS,
                              gov_exp_edu=SE.XPD.TOTL.GB.ZS,
                              f_ptime_chores=SG.TIM.UWRK.FE,
                              fertility_rate=SP.DYN.TFRT.IN)

wb_data <- wb_data %>% filter(!is.na(f_lfpr))

#split data into training and test data set 
train_data <- wb_data %>% filter(year < 2016)
test_data <- wb_data %>% filter(year > 2015)
```

## Background and Problem Statement
<div class="black">
Benefits: 

 - Women empowerment and bargaining power 
 - Domestic violence goes down when wage gap narrows 
 - Impact on children’s educational outcomes 
 - If male and female LFPR were equal by 2025, the annual global GDP would increase by 26% 

Problem:

 - Female LFPR and GDP per capita:U-shaped
 - Some countries: female LFPR is decreasing despite economic growth 
</div>

## Data 

<div class="black">

World Bank Database - the "wbstats" R-package

Data from Web Scraping

 - safety ranking
 - minimum wage law
 - religion
 - etc.

</div>

## Current Variables {.smaller}

<div class="black">
Dependent Variable: Female LFPR (interval-based outcome)

Independent Variables(8): 

  - GDP per capita
  - % literacy rate among adult female 
  - fertility rate (births per woman)
  - time spent on unpaid domestic and care work for female (% of 24 hour day)
  - compulsory education (years)
  - % government expenditure on education
  - % female aged 25+ completed at least lower secondary education
  - % female aged 25+ completed at least Bachelor's or equivalent
</div>

## Methods and Approaches

```{r}
p1<- ggplot(data=train_data) +
  geom_point(aes(x=gdp_pcap, y=f_lfpr,alpha=0.3))+
  theme_minimal()+
  theme(legend.position = "none",
        plot.title = element_text(face="bold",hjust=0.5))+
  labs(title = "Female LFPR vs GDP Per Capita",
       x="GDP Per Capita",
       y="Female LFPR")
```

```{r}
p2 <- ggplot(data=train_data)+
  geom_point(aes(x=fertility_rate,y=f_lfpr,alpha=0.8,color="red"))+
  theme_minimal()+
  theme(legend.position = "none",
        plot.title = element_text(face="bold",hjust=0.5))+
  labs(title = "Female LFPR vs Fertility Rate",
       x="Fertility Rate",
       y="Female LFPR")

```

```{r}
p3 <- ggplot(data=train_data,aes(x=f_lower_sec_edu,y=f_lfpr))+
  geom_point(color="darkgreen",alpha=0.8)+
  geom_smooth(method='lm',se=FALSE)+
  theme_minimal()+
  theme(legend.position = "none",
        plot.title = element_text(face="bold",hjust=0.5))+
  labs(title = "Female LFPR vs Secondary Education",
       x="% female aged 25+ completed at least lower secondary education",
       y="Female LFPR")
```

```{r}
p4 <- ggplot(data=train_data,aes(x=com_edu,y=f_lfpr))+
  geom_point(color="steelblue",alpha=0.8)+
  theme_minimal()+
  theme(legend.position = "none",
        plot.title = element_text(face="bold",hjust=0.5))+
  labs(title = "Female LFPR vs Compulsory Education",
       x="Years of Compulsory Education",
       y="Female LFPR")
```


```{r,warning=FALSE}
grid.arrange(p1,p2,p3,p4,nrow=2,ncol=2)
```

## Methods and Approaches

<div class="black">
General: Exploratory Data Analysis

Supervised Learning (regression):

 - K-Nearest Neighbors
 - Regression Trees
 - Random Forest 
</div>

## Analysis {.smaller}

```{r}
glimpse(train_data)
```

<div class="black">

Unit of Analysis: country-year

Problem: lots of missing values 

Solution: 

 - removed all obs with no info on female LFPR
 - imputed all missing values for every independent variable 
  
</div>

## Analysis {.smaller}

```{r,echo=TRUE}
#Reprocess the data
train_data1 <- train_data %>% select(-ccode,-country)

my_recipe <- recipe(f_lfpr~., data=train_data1) %>%
  step_knnimpute(all_predictors()) %>%
  step_range(all_numeric()) %>%
  prep()

train_data2 <- bake(my_recipe, train_data)
test_data2 <- bake(my_recipe, test_data)
```

```{r,echo=TRUE}
#Set cross-validation
set.seed(5678)
folds <- createFolds(train_data2$f_lfpr,k=5)
sapply(folds,length)
cross_val <- trainControl(method = "cv", index = folds)
```

## Analysis {.smaller}

```{r,echo=TRUE,results='hide'}
#K-nearest Neighbors(KNN)
knn_tune = expand.grid(k=c(1,5,10,50))

model_knn <- train(f_lfpr~.,data = train_data2, method="knn",
                   metric="RMSE",trControl=cross_val, 
                   tuneGrid=knn_tune)
```

```{r,echo=TRUE,results='hide'}
#Random Forest(ranger)

model_rf <- train(f_lfpr~., data=train_data2,method="ranger",
                  metric="RMSE", trControl=cross_val)

```

## Results

```{r,include=FALSE}
model_all <- list(knn=model_knn, rf=model_rf)
resamples(model_all)
p5 <- dotplot(resamples(model_all), metric= "RMSE")
```

```{r}
#Compare Models 
p5
```


## Results {.smaller}

```{r,echo=TRUE}
#Test the out-of-sample predictive performance
prediction <-predict(model_knn, newdata = test_data2)
rmse = sqrt(sum((test_data2$f_lfpr - prediction)^2/nrow(test_data2)))
rmse
```


## Lessons Learned and Next Steps 

<div class="black">

Lessons Learned:

  - Low RMSE (good prediction): indepedent variables capture the info related to education, economic development and demographic indicators 
  - Probably not the whole story: other factors that affect female LFPR

Next Steps:
  
  - Obtain more data 
  - Repeat EDA to understand the data better 
  - Estimate the female LFPR at the micro level 
  
</div>











